Isolator: accurate and consistent analysis of RNA-Seq experiments

### 2015-08-08 paper.tex
Stuff on qpcr validation.

The four SEQC samples consist of two commercially available reference RNA
samples, labeled $A$ and $B$, and two synthetic samples formed by mixing $A$ and
$B$ in specific proportions. Sample $C$ is composed of 75\% $A$ and 25\% $B$,
while sample $D$ is 25\% $A$ and 75\% $B$. This design allows us to directly
measure expression consistency. If we estimate transcript abundance for each
sample, producing estimates $a, b, c, d$, then for these estimates to be in
agreement we should expect $c$ to be approximated by $0.25a + 0.75b$, and
similarly $d$ by $0.75a + 0.25b$. To capture the notion of conistency, we
measured the closeness of this approximation using proportionality correlation
(Table \ref{table:seqctranscripts}).

Results from RSEM are particularly informative. We generated maximum
likeilood and posterior mean estimates using the same aligned reads.
Merely switching the estimate from the former to the latter increased
the correlation from 0.922 to 0.968. This stands in contrast to
comparisons to qPCR which suggest reduced accuracy of posterior mean
estimates, demonstrating that consistency and accuracy, while related,
are two separate axis of comparison.

@@

By modeling the entire RNA-Seq experiment and saving samples generated
during sampling, Isolator is uniquely able to compute posterior
probabilities corresponding to arbitrarily complex questions, within the
confines of the model. As a case study, we previously used Isolator to
analyze splicing dynamics in human cardiomyocyte cells during
maturation.

The experiment consisted of a number of conditions at different time
points so it did not easily lend itself to analysis using pairwise
hypothesis tests. We instead looked for genes showed a change in
splicing that was consistent between mature and immature cells by
computing a probability of ``monotonic splicing'', or specifically that
an observed change in splicing occurs in a consistent direction.

TODO: talk about validation.

>>>>

The four SEQC samples consist of two commercially available reference RNA
samples, labeled $A$ and $B$, and two synthetic samples formed by mixing $A$ and
$B$ in specific proportions. Sample $C$ is composed of 75\% $A$ and 25\% $B$,
while sample $D$ is 25\% $A$ and 75\% $B$. This design allows us to directly
measure expression consistency. If we estimate transcript abundance for each
sample, producing estimates $a, b, c, d$, then for these estimates to be in
agreement we should expect $c$ to be approximated by $0.25a + 0.75b$, and
similarly $d$ by $0.75a + 0.25b$. To capture the notion of consistency, we
measured the closeness of this approximation using proportionality correlation
(Table \ref{table:seqctranscripts}).

Results from RSEM are particularly informative. We generated maximum
likelihood and posterior mean estimates using the same aligned reads.
Merely switching the estimate from the former to the latter increased
the correlation from 0.922 to 0.968. This stands in contrast to
comparisons to qPCR which suggest reduced accuracy of posterior mean
estimates, demonstrating that consistency and accuracy, while related,
are two separate axis of comparison.

@@

Splicing has been known to play an important role in devoloping hearts. Specific
splicing changes during maturation have been shown to be conserved across
chicken and mouse, and influenced by expression of CELF and MBNL
\cite{Kalsotra:2008ke}. Recently, stress-induced switching between two mutually
exclusive isoforms of KHK has been shown to be driver of pathological cardiac
hypertrophy \cite{Mirtschink:2015iq}.

We previously used Isolator to analyze splicing dynamics in human cardiomyocyte
cells during maturation \cite{Kuppusamy:2015ey}. By modeling the entire RNA-Seq
experiment and saving samples generated during sampling, Isolator is uniquely
able to compute posterior probabilities corresponding to arbitrarily complex
questions, within the confines of the model. As this experiment consisted of
a number of samples from a number of conditions and time points, it did not
easily lend itself to analysis using pairwise tests.

We instead searched for differential splicing events by computing the
probability of ``monotonic splicing``, which we defined as a splicing change
that that occurs is consistently observed as occuring in the same direction
between immature (H7 cells at day 20 and at day 30 infected with an empty
vector) and mature samples (H7 1 Year, adult heart, and fetal atrium and
ventricle). To compute the posterior probability of monotonic splicing, we
simply count the number of samples from the model in which this pattern is
observed, and divide by the total number of samples generated (350, in this
case).

This approach identified 272 probable (posterior probability > 0.5) differential
splicing events. We chose seven high probability candidates (ACOT9, IMMT, LTBP4,
OBSL1, OSBPL1A, PALM, and TNNT2) to validate with qPCR (Supplementary Section
\ref{cardioqpcr}), by measuring the expression of specific exons predicted a
change in splicing. Five of these candidates (ACOT9, LTBP4, OBSL1, OSBPL1A, and
TNNT2) showed a clear gradient between H7 day20, fetal, and adult tissue.

We analyzed the same data with the other seven methods. Exon-level fold change
estimates generally agree between RNA-Seq methods, though far less so between
RNA-Seq and qPCR, which could be caused by poor gene annotations or yet
unidentified technical effects. Principal component analysis using transcript
expression estimates also demonstrates some revealing differences (Supplementary
Section \ref{cardiopca}).
